---
title: 'Reproducible Research: Peer Assessment 1'
output:
  html_document:
    keep_md: yes
    self_contained: no
---


## Loading and preprocessing the data
The data is assumed to be in a zip file in the same directory as this document. If it is not, download it. Unzip it and read it into a data frame.
```{r}
knitr::opts_chunk$set(fig.path='figure/')
options(scipen = 1, digits = 2)

# Check for the presence of the data, possibly zipped.
fileUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
zipfn <- "activity.zip"
csvfn <- "activity.csv"

# Make sure the data is available in the current directory.
if (!file.exists(csvfn)) {
    if (!file.exists(zipfn)) {
        # Download the data file
        download.file(fileUrl, zipfn, "curl")
    }
    # Unzip the data file
    unzip(zipfn)
}

data <- read.csv(csvfn)
```


## What is mean total number of steps taken per day?
Ignoring missing (NA) values, the number of steps taken per day is shown here:
```{r}
#cleanData <- subset(data, steps!="NA")
cleanData <- na.omit(data)
output1 <- aggregate(cleanData$steps, by = list(Date=cleanData$date), FUN=sum, simplify = TRUE)

library(ggplot2)
ticks <- c("2012-10-09","2012-10-16","2012-10-23","2012-10-30","2012-11-08","2012-11-18","2012-11-25")
ggplot(output1, aes(Date, x)) +
    geom_bar(stat="identity") +
    labs(title="Steps per Day") +
    ylab("Total Steps") +
    scale_x_discrete(breaks=ticks)

avgSteps <- mean(output1$x)
medSteps <- median(output1$x)
```
**The average number of steps per day is `r avgSteps`.**  
The median number of steps per day is `r medSteps`.  


## What is the average daily activity pattern?
Throughout the day the average number of steps varies as in this plot:
```{r}
output2 <- aggregate(cleanData$steps, by = list(Time=cleanData$interval), FUN=mean, simplify = TRUE)

plot(output2$Time, output2$x, main="Average Steps throughout the Day", xlab="Time", ylab="Average Steps", type="l")

maxInterval <- output2[which.max(output2$x),"Time"]
```
**The time interval with the greatest number of steps (on average over all days) is `r maxInterval`.**  


## Imputing missing values
```{r}
miss <- sum(is.na(data$steps))
```
The number of missing values in the data set is `r miss`.  
We will fill the missing values by using the average value for that time of day from the rest of the data set.  

```{r}
imputed <- data
imputed$steps[is.na(imputed$steps)] <- round(output2$x)

output3 <- aggregate(imputed$steps, by = list(Date=imputed$date), FUN=sum, simplify = TRUE)
ticks3 <- c("2012-10-07","2012-10-14","2012-10-21","2012-10-28","2012-11-04","2012-11-11","2012-11-18","2012-11-25")
ggplot(output3, aes(Date, x)) +
    geom_bar(stat="identity") +
    labs(title="Steps per Day") +
    ylab("Total Steps") +
    scale_x_discrete(breaks=ticks3)

avgSteps3 <- mean(output3$x)
medSteps3 <- median(output3$x)
```
**The average number of steps per day is `r avgSteps3`.**  
The median number of steps per day is `r medSteps3`.  

These values are slightly, but not significantly, lower than those calculated by skipping the missing data, so this method of imputing missing data seems like a good choice.

## Are there differences in activity patterns between weekdays and weekends?
In the following graph, we separate the steps by whether they occur on a weekday or a weekend.

```{r}
# Turn the date into a weekday/weekend factor
imputed$day <- as.factor(weekdays(as.Date(imputed$date),TRUE))
imputed$wkend <- as.factor(ifelse(grepl('^S', imputed$day),"weekend", "weekday"))

output4 <- aggregate(imputed$steps, by = list(Time=imputed$interval,Weekend=imputed$wkend), FUN=mean, simplify = TRUE)

library(lattice)
xyplot(x~Time|Weekend, data=output4, layout=c(1,2), type="l", ylab="Number of steps", xlab="Interval")
```

We can see that there are a number of differences:  

* The peak is around 8 AM on weekdays.
* Activity starts earlier on weekdays.
* Activity ends later on weekends.
* Activity is generally higher througout the day on weekends.


